---
title: "Benchmarking function calls in PhenotypeSimulator"
author: "Hannah Meyer"
date: "11/16/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(ggplot2)
library(microbenchmark)
```

Genotype simulation and covariance estimation are slow for large number of 
samples and SNPs. Find bottlenecks in PhenotypeSimulator v1.1 and update for 
next submission to CRAN and paper resubmission.

### simulateGenotypes()

```{r simulate-genotypes, cache=TRUE}
largeN <- 50000
smallN <- 100
largeNrSNP=50000
smallNrSNP=100
frequencies=c(0.1, 0.2, 0.4)
sampleID="ID_"
options(microbenchmark.unit="s")

bm_simulate <- microbenchmark(
    simulate_via_sapply_largeN_smallNrSNP = {
        samples <- paste(sampleID, seq(1, largeN, 1), sep="")
        freq <- sample(frequencies, smallNrSNP, replace=TRUE)
        X <- sapply(1:smallNrSNP, function(x) rbinom(largeN, 2, freq[x]))
    },
    simulate_via_matrix_largeN_smallNrSNP = {
        samples <- paste(sampleID, seq(1, largeN, 1), sep="")
        freq <- sample(frequencies, smallNrSNP, replace=TRUE)
        X <- matrix(rbinom(largeN*smallNrSNP, 2, prob=rep(freq, each=largeN)), 
                    nrow=largeN)
    },
    simulate_via_sapply_smallN_largeNrSNP = {
        samples <- paste(sampleID, seq(1, smallN, 1), sep="")
        freq <- sample(frequencies, largeNrSNP, replace=TRUE)
        X <- sapply(1:largeNrSNP, function(x) rbinom(smallN, 2, freq[x]))
    },
    simulate_via_matrix_smallN_largeNrSNP = {
        samples <- paste(sampleID, seq(1, smallN, 1), sep="")
        freq <- sample(frequencies, largeNrSNP, replace=TRUE)
        X <- matrix(rbinom(smallN*largeNrSNP, 2, prob=rep(freq, each=smallN)), 
                    nrow=smallN)
    },
    assign_SNPIDs_sampleNames_smallN_largeNrSNP = { 
        colnames(X) <- paste("SNP", 1:ncol(X), sep="")
        rownames(X) <- samples
    },
    assign_SNPIDs_sampleNames_largeN_smallNrSNP = { 
        colnames(X) <- paste("SNP", 1:ncol(X), sep="")
        rownames(X) <- samples
    },
    times = 10
)

autoplot(bm_simulate)
```

```{r simulate-summary, cache=TRUE, echo=FALSE }
kable(summary(bm_simulate, unit="s"), digits = 2, 
      caption="Execution times of genotype simulation and ID assignment")
```



### getAlleleFrequencies()
Genotype counting initially based on as.data.frame(table(snp)). Changed to 
specifically counting 0,1,2 and raising an error when other genotype codes are
supplied. 

```{r counts, cache=TRUE }
options(microbenchmark.unit="relative")
snp <- rbinom(100000, 2, prob=0.2)

bm_counts <- microbenchmark(
    count_via_table = {counts <- data.frame(table(factor(snp, 
                                                levels=c(0,1,2))))[,2]},
    count_via_length = {    pp <- length(which(snp == 0))
                  pq <- length(which(snp == 1))
                  qq <- length(which(snp == 2))
        
    },
    times=10
)

autoplot(bm_counts)
```


```{r counts-summary, cache=TRUE, echo=FALSE }
kable(summary(bm_counts, unit="relative"), digits = 2, 
      caption="Relative execution times of genotype counting")
```


### getKinship()

```{r covariance, cache=TRUE }
options(microbenchmark.unit="relative")
bm_covariance <- microbenchmark(
    cov_via_tcrossprod = { kinship <- tcrossprod(X) },
    cov_via_XXT = { kinship <-  X %*% t(X)},
    times = 10
)

autoplot(bm_covariance)
```


```{r covariance-summary, cache=TRUE, echo=FALSE }
kable(summary(bm_covariance), digits = 2, 
      caption="Relative execution times of genotype counting")
```



